Wrangling multivariate spatio-temporal data with the R package cubble

### 2022-04-27 1-intro.Rmd
misc small changes

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously. 
<!-- Underlying all these tasks, there is a data representation that enables the spatio-temporal data to be sliced and diced differently.  -->
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.**

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full di layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link]. **At this point what is cubble is adding with respect the other packages it is not  clear to me**

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation and a temporal component that is recorded at regular or irregular time intervals. It may also include multiple variables measured at each spatial and temporal values. With spatio-temporal data, one can fix the time to explore the spatial features of the data, fix the spatial location/s to explore temporal aspects, or dynamically explore the space and time simultaneously. 
<!-- Underlying all these tasks, there is a data representation that enables the spatio-temporal data to be sliced and diced differently.  -->
In order to computationally explore the spatial, temporal and spatio-temporal faces of such data, the data needs to be stored and represented under a specific data object that allows the user to query, group and dissect all the data faces.

@@

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately or combined while being kept synchronised. Among the four spacetime layouts in @spacetime, \pkg{cubble} can be applied to full di layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at https://CRAN.R-project.org/package=cubble. **At this point what is cubble is adding with respect the other packages it is not  clear to me**

### 2022-04-27 2-0-cube.Rmd
misc small changes

\textcolor{blue}{Ursula: I have rearranged this section somewhat, please check!}

@@

The most common data cube using the three axes (time, latitude, longitude)  could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the workhorse behind \pkg{cubble}.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array,  may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) The `mutate()` syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.
\textcolor{blue}{Ursula: this seems too specific, you should be able to make the same argument without referencing the mutate function?}
\textcolor{green}{How about the other functions from dplyr?, the same principle would apply right?}

>>>>

@@

The most common data cube using the three axes (time, latitude, longitude)  could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind \pkg{cubble}.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array,  may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) Creating new variables in an array is more complex than in a data frame since it now needs to be defined in a 2D space, rather than a 1D vector, to be added into the array.

### 2022-04-27 3-advanced.Rmd
fix merge conflict
misc small changes

## Data fusion and matching

@@

In \proglang{R}, \pkg{GGally} implements glyph maps through the \code{glyphs()} function. The function constructs a data frame with calculated position (\code{gx}, \code{gy}. \code{gid}) of each point on the time series using linear algebra (Equation 1 and 2 in @Wickham2012-yr). The data can then be piped into \code{ggplot} to create the glyph map as:

>>>>

## Data fusion and matching {#matching}

@@

In \proglang{R}, \pkg{GGally} implements glyph maps through the \code{glyphs()} function. The function constructs a data frame with calculated position (\code{gx}, \code{gy}, \code{gid}) of each point on the time series using linear algebra (Equation 1 and 2 in @Wickham2012-yr). The data can then be piped into \code{ggplot} to create the glyph map as:

### 2022-04-27 4-1-basic.Rmd
misc small changes

```{r}
load(here::here("data/covid.rda"))
lga <- absmapsdata::lga2018 %>%
  rename(lga = lga_name_2018) %>%
  filter(state_name_2016 == "Victoria") %>%
  rmapshaper::ms_simplify()
```

>>>>

```{r}
load(here::here("data/covid.rda"))

lga <- strayr::read_absmap("lga2018") %>%
  rename(lga = lga_name_2018) %>%
  filter(state_name_2016 == "Victoria") 
```

### 2022-04-27 5-conclusion.Rmd
misc small changes

This paper describes an \proglang{R} package \pkg{cubble} for manipulating and visualising spatio-temporal data. A new data structure, \code{cubble} that builds from the \code{rowwise_df} and \code{grouped_df} class in the tidyverse ecosystem, is proposed to connect the time invariant and varying variables in the spatio-temporal data. This design frees the data analysts from spending time on organising variables of different observational units. The data structure is also flexible to the techniques and packages analysts use to analyse the data, for example, in the matching example in section 4.3, users are free to use algorithms from another package to cluster stations.

Further development and maintenance of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. In the spatial aspect, the simple feature representation wraps spatial coordinates in a list column, for some complex spatial operations, while sometimes, analysts may want to unpack them into longitude and latitude columns to work from. Another extension to \pkg{cubble} is to build a smooth transition between the coordinate columns and simple feature geometry column.

>>>>

This paper describes an \proglang{R} package \pkg{cubble} for organising, manipulating and visualising spatio-temporal data. A new data structure, \code{cubble}, with the nested and long form, is proposed to connect the time invariant and varying variables in the spatio-temporal data. This design frees the data analysts from spending time on organising variables of different observational units. The data structure is also flexible to the techniques and packages analysts use to analyse the data, for example, in the matching example in section 4.3, users are free to use algorithms from another package to cluster stations.

Future directions of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. In the spatial aspect, the simple feature representation wraps spatial coordinates in a list column, for some complex spatial operations, while sometimes, analysts may want to unpack them into longitude and latitude columns to work from. Another extension to \pkg{cubble} is to build a smooth transition between the coordinate columns and simple feature geometry column.

### 2022-04-27 arxiv/cubble-arxiv.Rmd
misc small changes

---
title: "cubble: An R Package for Structuring Spatio-temporal Data"
authors:
  - name: H. Sherry Zhang
    affiliation: Monash University
    location: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    location: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    location: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: BNU-HKBU United International College
    location: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: Monash University
    location: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
abstract: |
    Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyse different aspects of the spatio-temporal data simultaneouly, like for instance instance, temporal trends of multiple variables across locations. In order to facilitate the study of different sections or combinations of spatio-temporal data, we introduce a new data structure, \code{cubble}, with a suite of functions enabling easy slicing and dicing on the different components spatio-temporal components, for visual and numerical explorations while facilitating data wrangling for modelling. The propose \code{cubble} structure ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. The \code{cubble} structure and the tools provided in the \code{cubble} R package equip users with the  capability to handle hierarchical spatial and temporal structures. \color{green}{The tools are illustrated with examples of Australian climate data, merging climate and river level data sources, reproducing a climate reanalysis (ERA5), and creating interactive graphics.}
keywords:
- spatial, temporal, spatio temporal, R, exploratory data analysis, environmental data, climate data
bibliography: "../references.bib"
cls: "../jss.cls"
bst: "../jss.bst"
preamble: |
  \newcommand{\sherry}[1]{\textcolor{magenta}{#1}}
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
link-citations: true
---
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = here::here("figures/"),
  fig.height = 10,
  root.dir = here::here()
)
options(prompt = "R> ", continue = "+ ",
        tibble.print_max = 5, tibble.print_min = 5)
```

>>>>

---
title: "Wrangling multivariate spatio-temporal data with the R package cubble"
authors:
  - name: H. Sherry Zhang
    affiliation: Monash University
    location: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    location: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    location: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: BNU-HKBU United International College
    location: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolaslangrene@uic.edu.cn
  - name: Patricia Menéndez
    affiliation: Monash University
    location: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
abstract: |
    Multivariate spatio-temporal data refers to multiple measurements taken across space and time. For many analyses, spatial and time components can be separately studied: for example, to explore the temporal trend of one variable for a single spatial location, or to model the spatial distribution of one variable at a given time. However for some studies, it is important to analyse different aspects of the spatio-temporal data simultaneouly, like for instance, temporal trends of multiple variables across locations. In order to facilitate the study of different sections or combinations of spatio-temporal data, we introduce a new data structure, \code{cubble}, with a suite of functions enabling easy slicing and dicing on the different components spatio-temporal components, for visual and numerical explorations while facilitating data wrangling for modelling. The proposed \code{cubble} structure ensures that all the components of the data are easy to access and manipulate while providing flexibility for data analysis. The \code{cubble} structure and the tools provided in the \code{cubble} R package equip users with the capability to handle hierarchical spatial and temporal structures. The tools are illustrated with examples of Australian climate data, merging climate and river level data sources, reproducing a climate reanalysis (ERA5), and creating interactive graphics.
keywords:
- spatial, temporal, spatio temporal, R, exploratory data analysis, environmental data, climate data
bibliography: "../references.bib"
cls: "../jss.cls"
bst: "../jss.bst"
preamble: |
  \newcommand{\sherry}[1]{\textcolor{magenta}{#1}}
output:
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
link-citations: true
---
```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  fig.path = here::here("figures/"),
  fig.height = 10,
  root.dir = here::here()
)
options(prompt = "R> ", continue = "+ ",
        tibble.print_max = 5, tibble.print_min = 5)
```
